---
project:
  type: website

website:
  title: "CIS9660 Lab Compilation"
  author: "Chris Panlasigui, Yuanfeng Cai"
format: 
  html:
    code-tools: true
    toc: true
    toc-depth: 3
    code-fold: show
    code-summary: "Code"
    code-overflow: wrap 
editor: visual
execute:
  output: false
---


# Exploratory Data Analysis {#sec-1eda}

## Loading Data {#sec-loaddata}

Before the lab, please download and save data `AutoLab1.csv` from Blackboard-Data.

```{r}
# use file.choose() if you want a window to open for you to locate your file
# Auto=read.csv(file.choose(),header=T, stringsAsFactors=TRUE)

# if you know the path to your file you can directly include it and omit file.choose()
Auto=read.csv(file="../raw_data/autolab1.csv",
              header=T, 
              stringsAsFactors=TRUE)
```

You can use function `read.table()` or `read.csv()` to import data into R, depending on the file type. You can use `help(read.csv)` or `?read.csv` to find more about how to use this function. Since the file type is .csv, we use `read.csv()`in this case. Use the following command to load `AutoLab1.csv` into R and store it as an object called Auto.

The command, `file.choose()` locates the file from your local drive, the option `header=T` (or `header=TRUE`) means that the first line of the file contains the variable names, the option `stringsAsFactors=TRUE` (or `stringsAsFactors=T`) means that converting characters to factors

## Data Structure {#sec-1datastruct}

```{r}
head(Auto)
dim(Auto)
str(Auto)
```

You can use `head()` to look at the first few rows. Function `dim()` can output the number of rows followed by the number of columns and function `str()` can return the data structure.

## Summary Statistics {#sec-1summarystat}

```{r}
summary(Auto)
```

We can use function `summary()` to output the summary statistics of data.

## Data Restructuring {#sec-1datarestruct}

```{r}
Auto[,2]=as.factor(Auto[,2])
summary(Auto[,2])
```

The variable cylinders (i.e. the second column) is stored as a quantitative variable. As it only has a small number of possible values, we can convert it to a qualitative variable. Function as.factor() converts a quantitative variable into a qualitative variable. Check the summary statistics of cylinders to see if it is the same as the output in @sec-1summarystat.

## Graphs {#sec-1edagraphs}

### Scatterplot {#sec-1edascatter}

```{r}
plot(x=Auto$horsepower, y=Auto$mpg)
```

We can use function plot() to produce plots. However, simply typing the variable names does not work, because R does not know where to find those variables. You can either use the `$`sign or 

```{r}
attach(Auto)
names(Auto)
plot(x=horsepower , 
     y=mpg, col ="red", 
     xlab="Horsepower",
     ylab ="MPG ",
     xlim=c(30,250), 
     ylim=c(5,50), 
     main="Horsepower vs. MPG")
```

Use function `attach()` to tell R to make the variables available by name. Function `names()` lists all variable names. Then you can use the variable name directly.

Here the option `col ="red"` tells R that color data points red. The options `xlab="Horsepower",ylab ="MPG"`, and `main="Horsepower vs. MPG"` tell R the x axis title, the y axis title and the main title respectively.

The options `xlim` and `ylim` tell R the range of x axis and y axis.
There are many other optional parameters in function `plot()`, which we do not include in this case. You can use `help(plot)` or `?plot` to explore more about them.

```{r}
par(mfrow=c(1,2))
plot(x=acceleration ,
     y=mpg, 
     col ="red", 
     xlab="Acceleration",
     ylab ="MPG ", 
     main="Acceleration vs. MPG")

plot(x=weight , 
     y=mpg, 
     col ="red", 
     xlab="Weight", 
     ylab ="MPG ", 
     main="Weight vs. MPG")
```

We can use function `par(mfrow=c(nrows,ncols))` to combine multiple plots into one graph. For example, `par(mfrow=c(3,1))` indicates that three figures will be arranged in 3 rows and 1 column. Now let’s generate another two figures and arrange them in one column.

```{r}
pairs(Auto)
pairs(Auto[c(3:5)])
```

The `pairs()` function creates a scatterplot for every scatterplot pair of variables. We can also produce scatterplots matrix for just a subset of the variables.


### Barplot {#sec-edabarplot}

```{r}
par(mfrow=c(1,1))
plot(x=cylinders, 
    y=mpg, 
    col="red", 
    varwidth=T, 
    xlab=" Cylinders ", 
    ylab ="MPG ", 
    main="Cylinders vs. MPG")
```

If the variable plotted on the x-axis is categorical, then boxplots will automatically be produced.

### Histogram {#sec-edahist}

```{r}
hist(x=mpg, 
     breaks =10, 
     col="red", 
     xlab="MPG ",
     xlim=c(0,50),
     main="Histogram of MPG")
```

We can use `hist()` function to plot a histogram. The option `“breaks=10”` sets the total number of bins.


# Linear Regression {#sec-2linearregression}

## Load Data {#sec-2loaddata}

Before the lab, please download Data Lab2Data.csv from Blackboard→Data. Then load Data Lab2Data.csv to object Datlab2.

```{r}
# delete global environment
rm(list = ls())

# load data
Datlab2 = read.csv(file="../raw_data/Lab2Data.csv", 
                   header=T, 
                   stringsAsFactors=TRUE)

# check column/variable names
names(Datlab2)

attach(Datlab2)
```

There are four input variables $(𝑥_1 ...𝑥_4)$ and one response variable $y$ in this data. Variable $x_4$ is a categorical variable with two possible values. We should create a dummy variable that takes on two possible numerical values. 

```{r}
contrasts(x4)
```

The `contrasts()` function returns how R codes this dummy variable.

## Build Model {#sec-2buildmodel}

### Full Model {#sec-2fullmodel}

```{r}
Model1=lm(y~x1+x2+x3+x4,data= Datlab2)
```

We first fit the entire data with all input variables into a multiple linear regression model Model1. We can use the `lm()` function:

$$ y =\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_4 + \epsilon$$ {#eq-linearm1}

```{r}
summary(Model1)
```

We can use the function `summary()` to obtain the coefficients, their associated `p-values`, and the $R^2$. We can find that only one variable looks insignificant in this model.

### Model with interaction effect {#sec-2interactioneffect}

```{r}
Model2=lm(y~x3*x4)
summary(Model2)
```


We can also check if the model should include an interaction effect. For example, to test an interaction effect between $x_3$ and $x_4$, we can include `lm(y~x3*x4)`. The syntax `x3*x4` in the function `lm()` simultaneously includes $x_3$, $x_4$, and $x_3 \times x_4$𝑥4. That is, `lm(y~x3*x4)` fits the following model:

$$y = \beta_0 + \beta_1 x_3 + \beta_2 x_4 + \beta_3(x_3 \times x_4) + \epsilon$$ {#eq-linearm2}
The result suggests that the interaction effect is significant.

```{r}
plot(x=predict(Model2), y=residuals(Model2))
plot(x=predict(Model2), y=rstudent(Model2))
```

We can generate diagnostic plots for models. The function `rstudent()` returns the studentized residuals, which can be used to identity outliers. Given the diagnostics results, there is no outlier in `Model2`.


## Resampling Methods {#sec-2resampling}

Now we try two candidate models `M1` and `M2`.  We will use re-sampling methods to compare their performance. 

$$M1: y =\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_4 + \beta_5(x_3 \times x_4) +  \epsilon$$ {#eq-M1}

$$M2: y = \beta_0 + \beta_1 x_3 + \beta_2 x_4 + \beta_3(x_3 \times x_4) + \epsilon$$ {#eq-M2}


### Hold-out {#sec-2holdout}

We use two types of re-sampling methods. First, we illustrate how to compare the models using hold-out. We use 80-20 hold-out in this case.  We develop models using a training data set and assess the model performance using a test data set. 

```{r}
# reproduceable random sampling
set.seed(1)

#index for training data set
train=sample(nrow(Datlab2),nrow(Datlab2)*0.8) 

#training data set
Datlab2.train=Datlab2[train, ] 

#test data set
Datlab2.test=Datlab2[-train, ] 

#the response in the test set
y.test=y[-train] 
```

We first use the `sample()` function to randomly split the original data set into one training set and one test set.

The function `nrow()` counts the total number of rows in `Datlab2`. Sometimes we want to reproduce the exact same sampling results; we can use the `set.seed()` function. To use this function, you need to set a seed, which is an arbitrary integer, e.g. 1.

```{r}
# full model with train data
M1train=lm(y~x1+x2+x3*x4, data=Datlab2.train)

# interaction model with train data
M2train =lm(y~x1+x3*x4, data=Datlab2.train)
```

The first model of choice `M1` contains all four variables and one interaction term. The second model `M2` contains three input variables and one interaction term. We learn the two models only using the training data set.

```{r}
# test performance of full model
y.predictM1=predict(M1train,Datlab2.test)
# performance measure using MSE of full model
M1MSE=mean((y.test-y.predictM1)^2)

# test performance of interaction model
y.predictM2=predict(M2train,Datlab2.test)
# performance measure using MSE of interaction model
M2MSE=mean((y.test-y.predictM2)^2)
```

To compare the model performance, we calculate the MSE of each model using the testing data set. The function `predict()` predicts the value of the response variable for each observation in test data. `y.predictM1` stores the predicted value of each observation in the test data set using model `M1`, and `y.predictM2` stores the predicted value of each observation in the test data set using model `M2`. The `mean()` function here is to calculate the average prediction deviation, i.e. MSE, in the entire test data.

#### AIC, BIC, Adjusted $R^2$ 


```{r}
# other performance measure of full model
AIC(M1train) # AIC
BIC(M1train) # BIC
summary(M1train) #adj R2

# other performance measure of interaction model
AIC(M2train) # AIC
BIC(M2train) # BIC 
summary(M2train) # adj. R2
```

In addition to using MSE, the model can be assessed using functions `BIC()`, `AIC()` and adjusted R2 in `summary()`.

### Cross-Validation (CV)

Now we illustrate how to compare the models using cross-validation. We use 5-fold cross validation along with MSE in this case. 

```{r}
# reproduceable random sampling
set.seed(1) 

# number of folds
k=5

# empty container to hold MSE of full model using CV
M1CVMSE=rep(0,k)
# emtpy container to hold MSE of interaction model using CV
M2CVMSE=rep(0,k)
```

We first create a vector to store the accuracy results associated with each fold for each model. We set the initial values for this vector as zero.

```{r}
#| output: true
#| include: true

# split data into k=5 folds 
folds=sample(1:k,nrow(Datlab2),replace=TRUE)
folds
```

We use the `sample()` function to split the original data set into five folds. This creates a vector of random values between 1 through 5.  This assigns the indices of `Datlab2` to each fold.  

```{r}
for(j in 1:k) # iterate from 1 through k=5
{ 
  # model of the train data portion of cv 
  M1CV=lm(y~ x1+x2+x3*x4,data=Datlab2[folds!=j,]) 
  # obtain the performance (MSE) of the model M1CV by applying it to the test portion of cv  
  M1CVMSE[j]=mean((y-predict(M1CV,Datlab2))[folds==j]^2) }

for(j in 1:k)
{ M2CV=lm(y~ x1 +x3*x4,data=Datlab2[folds!=j,]) 
M2CVMSE[j]=mean((y-predict(M2CV,Datlab2))[folds==j]^2) }
```

During each iteration, the `Datlab2[folds!=j,]` selects indices that are not equal to the $jth$ iteration and assigns it as the train data and stores the model value in `M1CV`.  So during the first iteration, the data train data will be all the indices that are not assigned to 1 in the `folds` vector. 

`[folds==j]` selects the indices that are equal to the $jth$ iteration, which is the test data.  So, `(y-predict(M1CV,Datlab2))[folds==j]^2` is the standard error of the test data, which the difference between the actual values of `y` to the predicted values based on the model `M1CV`.  MSE is calculated by obtaining the mean of the standard error in each iteration and is stored in `M2CVMSE[j]` where `j` is the current iteration index. `M2CVMSE` will then have 5 values since `j` has been set from 1 through `k=5`.

```{r}
MeanM1MSE=mean(M1CVMSE) ###CVMSE-M1###
MeanM2MSE=mean(M2CVMSE) ###CVMSE-M2###
```

Finally, calculate the cross-validation MSE of `M1` (@eq-M1) and `M2` (@eq-M2) by obtaining the mean of each CV model. 

# Classification: Logistic Regression & KNN {#sec-3class1}

## Logistic Regression {#sec-3logistic}

```{r}
#| output: true
#| 
# clear global env (remove all stored obj) 
rm(list = ls())

# load data
Default=read.csv(file = "../raw_data/Default.csv",
                 header=T, stringsAsFactors=TRUE)
attach(Default)

# check the dimensions
dim(Default)
names(Default)
```

Load Data `Default.csv` to object Default. The original data set contains 10000 observations.

```{r}
glm.fit1=glm(default~balance+income+student,
             family="binomial",
             data=Default)
```

First, we can use the `glm()` function to fit model `glm.fit1` which includes all three independent variables: `student`, `balance` and `income`, with all the observations. 

```{r}
summary(glm.fit1)
coef(glm.fit1)
exp(coef(glm.fit1))
```

We can use the function `summary()` to obtain the coefficients, their associated p-values, model AIC and residual deviance. The function `exp()` calculates **odds ratio** of the coefficients.

### Prediction accurary {#sec-3accuracy}

```{r}
set.seed(1)

# 80/20 hold-out re-sampling
# indices of 80% train data 
train=sample(nrow(Default),nrow(Default)*0.8)

# indices 20% test data
Default.test=Default[-train, ]

# y-values (default variable) of the 20% test data
test.truevalue=default[-train]
```

We next evaluate the prediction accuracy of this model. We use the `sample()` function to split the original data set into one training set and one test set. We randomly select 80% of the observations for training and
the remaining in the test set, `Default.test`. And save the true value of the testing set in the vector `test.truevalue`.


```{r}
glm.fit2=glm(default~balance+student+income,
             data=Default,
             subset=train,
             family =binomial)
```

We fit the data with all three independent variables into a logistic regression model `glm.fit2`. The *`subset=train`* option in `glm()` is to fit a regression using only the observations corresponding to the subset.


```{r}
glm.probs2=predict(glm.fit2,Default.test, type="response")
```

We evaluate the performance of the model using the test set (Default.test). We use the function `predict()` to calculate the *predicted probabilities* of the **default** in the test set and store them to `glm.pred2`. The `type="response"` option tells R to output probabilities not the logit.



















# Classification: CART, Bagging & Random Forest {#sec-4class2}

asdf
