---
project:
  type: website

website:
  title: "CIS9660 Lab Compilation"
  author: "Chris Panlasigui, Yuanfeng Cai"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show the code"
editor: visual
execute:
  output: false
---

# Lab Compilation

## Exploratory Data Analysis {#sec-EDA}

### Loading Data {#sec-loaddata}

Before the lab, please download and save data `AutoLab1.csv` from Blackboard-Data.

```{r}
# use file.choose() if you want a window to open for you to locate your file
# Auto=read.csv(file.choose(),header=T, stringsAsFactors=TRUE)

# if you know the path to your file you can directly include it and omit file.choose()
Auto=read.csv(file="../raw_data/autolab1.csv",
              header=T, 
              stringsAsFactors=TRUE)
```

You can use function `read.table()` or `read.csv()` to import data into R, depending on the file type. You can use `help(read.csv)` or `?read.csv` to find more about how to use this function. Since the file type is .csv, we use `read.csv()`in this case. Use the following command to load `AutoLab1.csv` into R and store it as an object called Auto.

The command, `file.choose()` locates the file from your local drive, the option `header=T` (or `header=TRUE`) means that the first line of the file contains the variable names, the option `stringsAsFactors=TRUE` (or `stringsAsFactors=T`) means that converting characters to factors

### Data Structure {#sec-datastruct}

```{r}
head(Auto)
dim(Auto)
str(Auto)
```

You can use `head()` to look at the first few rows. Function `dim()` can output the number of rows followed by the number of columns and function `str()` can return the data structure.

### Summary Statistics {#sec-summarystat}

```{r}
summary(Auto)
```

We can use function `summary()` to output the summary statistics of data.

### Data Restructuring {#sec-datarestruct}

```{r}
Auto[,2] =as.factor (Auto[,2])
summary(Auto[,2])
```

The variable cylinders (i.e. the second column) is stored as a quantitative variable. As it only has a small number of possible values, we can convert it to a qualitative variable. Function as.factor() converts a quantitative variable into a qualitative variable. Check the summary statistics of cylinders to see if it is the same as the output in @sec-summarystat.

### Graphs {#sec-edagraphs}

#### Scatterplot {#sec-edascatter}

```{r}
plot(x=Auto$horsepower , y=Auto$mpg)
```

We can use function plot() to produce plots. However, simply typing the variable names does not work, because R does not know where to find those variables. You can either use the `$`sign or 

```{r}
attach(Auto)
names(Auto)
plot(x=horsepower , 
     y=mpg, col ="red", 
     xlab="Horsepower",
     ylab ="MPG ",
     xlim=c(30,250), 
     ylim=c(5,50), 
     main="Horsepower vs. MPG")
```


Use function `attach()` to tell R to make the variables available by name. Function `names()` lists all variable names. Then you can use the variable name directly.


Here the option `col ="red"` tells R that color data points red. The options `xlab="Horsepower",ylab ="MPG"`, and `main="Horsepower vs. MPG"` tell R the x axis title, the y axis title and the main title respectively.

The options `xlim` and `ylim` tell R the range of x axis and y axis.
There are many other optional parameters in function `plot()`, which we do not include in this case. You can use `help(plot)` or `?plot` to explore more about them.

```{r}
par(mfrow=c(1,2))
plot(x=acceleration ,
     y=mpg, 
     col ="red", 
     xlab="Acceleration",
     ylab ="MPG ", 
     main="Acceleration vs. MPG")

plot(x=weight , 
     y=mpg, 
     col ="red", 
     xlab="Weight", 
     ylab ="MPG ", 
     main="Weight vs. MPG")
```

We can use function `par(mfrow=c(nrows,ncols))` to combine multiple plots into one graph. For example, `par(mfrow=c(3,1))` indicates that three figures will be arranged in 3 rows and 1 column. Now let’s generate another two figures and arrange them in one column.

```{r}
pairs(Auto)
pairs(Auto[c(3:5)])
```

The `pairs()` function creates a scatterplot for every scatterplot pair of variables. We can also produce scatterplots matrix for just a subset of the variables.


#### Barplot {#sec-edabarplot}

```{r}
par(mfrow=c(1,1))
plot(x=cylinders, 
    y=mpg, 
    col="red", 
    varwidth=T, 
    xlab=" Cylinders ", 
    ylab ="MPG ", 
    main="Cylinders vs. MPG")
```

If the variable plotted on the x-axis is categorical, then boxplots will automatically be produced.

#### Histogram {#sec-edahist}

```{r}
hist(x=mpg, 
     breaks =10, 
     col ="red", 
     xlab ="MPG ",
     xlim=c(0,50),
     main="Histogram of MPG")
```

We can use `hist()` function to plot a histogram. The option `“breaks=10”` sets the total number of bins.


## Linear Regression {#sec-linearregression}

### Load Data

Before the lab, please download Data Lab2Data.csv from Blackboard→Data. Then load Data Lab2Data.csv to object Datlab2.

```{r}
# delete global environment
rm(list = ls())

# load data
Datlab2 = read.csv(file="../raw_data/Lab2Data.csv", 
                   header=T, 
                   stringsAsFactors=TRUE)

# check column/variable names
names(Datlab2)

attach(Datlab2)
```

There are four input variables $(𝑥_1 ...𝑥_4)$ and one response variable $y$ in this data. Variable $x_4$ is a categorical variable with two possible values. We should create a dummy variable that takes on two possible numerical values. 

```{r}
contrasts(x4)
```

The `contrasts()` function returns how R codes this dummy variable.

## Build Model

### Full Model

```{r}
Model1=lm(y~x1+x2+x3+x4,data= Datlab2)
```

We first fit the entire data with all input variables into a multiple linear regression model Model1. We can use the `lm()` function:

$$ y =\beta_0 + \beta_1 x_1 + \beta_2 x_2 + \beta_3 x_3 + \beta_4 x_4 + \epsilon$$ 

```{r}
summary(Model1)
```

We can use the function `summary()` to obtain the coefficients, their associated `p-values`, and the $R^2$. We can find that only one variable looks insignificant in this model.

### Model with interaction effect

```{r}
Model2=lm(y~x3*x4)
summary(Model2)
```


We can also check if the model should include an interaction effect. For example, to test an interaction effect between $x_3$ and $x_4$, we can include `lm(y~x3*x4)`. The syntax `x_3*x_4` in the function `lm()` simultaneously includes $x_3$, $x_4$𝑥3, and $x_3 \times x_4$𝑥4. That is, `lm(y~x3*x4)` fits the following model:

$$y = \beta_0 + \beta_1 x_3 + \beta_2 x_4 + \beta_3(x_3 \times x_4) + \epsilon$$
