---
title: "CIS9660 Lab Compilation"
author: "Chris Panlasigui, Yuanfeng Cai"
format: 
  html:
    toc: true
    toc-depth: 2
editor: visual
---

# Exploratory Data Analysis

## Load Data

Before the lab, please download and save data `AutoLab1.csv` from Blackboard-Data.

You can use function `read.table()` or `read.csv()` to import data into R, depending on the file type. You can use `help(read.csv)` or `?read.csv` to find more about how to use this function.

Since the file type is .csv, we use `read.csv()`in this case. Use the following command to load `AutoLab1.csv` into R and store it as an object called Auto.

The command, `file.choose()` locates the file from your local drive, the option `header=T` (or

`header=TRUE`) means that the first line of the file contains the variable names, the option

`stringsAsFactors=TRUE` (or `stringsAsFactors=T`) means that converting characters to factors

```{r}
# use file.choose() if you want a window to open for you to locate your file
# Auto=read.csv(file.choose(),header=T, stringsAsFactors=TRUE)

# if you know the path to your file you can directly include it and omit file.choose()
Auto=read.csv("../raw_data/autolab1.csv",header=T, stringsAsFactors=TRUE)
```

## Data Structure

You can use `head()` to look at the first few rows. Function `dim()` can output the number of rows followed by the number of columns and function `str(auto)` can return the data structure.

```{r}
head(Auto)
dim(Auto)
str(Auto)
```

## Summary Statistics

We can use function `summary()` to output the summary statistics of data.

```{r}
summary(Auto)
```

## Data Restructuring

The variable cylinders (i.e. the second column) is stored as a quantitative variable. As it only has a small number of possible values, we can convert it to a qualitative variable. Function as.factor() converts a quantitative variable into a qualitative variable. Check the summary statistics of cylinders to see if it is the same as the output in Step 3.
